---
title: "Debugging in R"
description: "Tools and tips"
author: "Núria Estanyol"
date: "2023-06-21"
date-modified: last-modified
categories: 
  - debugging
  - error messages
  - namespace issues
  - function documentation
---

> This session was recorded and uploaded on YouTube here:

{{< video https://youtube.com/embed/eZtXI2hSZXI >}}

Debugging is a very broad topic, encompassing a wide array of scenarios
that are often highly context-specific. We can encounter diverse
challenges that need to adopt different debugging methods depending on
the problems we encounter.

Here we illustrate some of the fundamental steps to understanding bugs
and address some of the most common debugging strategies.

## Error messages

The initial step in debugging is to understand the error message. Error
messages are often quite explicit, offering hints of the type of errors,
locations, and suggestions to fix the issues. But they sometimes can be
*too* specific and get too technical so that it is hard to understand
what they mean.

Don't be alarmed by error messages! Encountering them is very common and
a regular part of the process no matter your skill level. Getting used
to the presence of error messages and learning to read them is
essential. Indeed, the very first step of debugging is to read the error
message. Let's begin with a straightforward example:

```{r}
#| error: true
sum(c("a","b"))
```

The `c()` function aggregates elements into a vector and the `sum()`
function computes the total from a vector of numeric values. As we can't
calculate the sum of letters, this leads to an issue. The prompted error
message provides a clue saying this is an *invalid "type"* assertion,
suggesting a character input.

## Reread your code

The second step of debugging involves rereading your code when
encountering an error.

In many instances, a significant part of the issues we face during
coding is overlooking a comma, a bracket, a quote, or other typo. In the
code below as an example, we forgot a comma. Unfortunately, the error
message is not very helpful in this case:

```{r}
#| error: true
paste("Hi!" "what's up?")
```

This is why it is important to read over the code again if you don't
understand the error message.

## Function documentation

In the previous example, if we understand the functionality of the
`paste()` function, we can identify the source of the error.

We can access the function documentation (see image below) by either
clicking on the function name and pressing `F1`, or by typing `?paste`
or `help(paste)` directly into the console.

![Documentation for the `paste()`
function.](images/help_documentation.png){#image-help-documentation
align="center"}

## Namespace issues

Namespace is the term used to describe what functions are exposed to
users inside a package. So the namespace of the package, for instance,
`{dplyr}`, includes the functions `select()`, `filter()`, or `mutate()`.
The functions and objects available in a package's namespace can be
found by using the `::` before the package name and hitting Tab to see a
list of objects (e.g. `dplyr::`).

Namespace issues occur when we want to use a function from a specific
package, but we're actually using a function from a different package.
For instance, consider the case of attempting to filter values within
the CO2 dataset. We might assume we are using the `filter()` function
from the `{dplyr}` package, so we run this:

```{r}
#| error: true
library(magrittr)
CO2 %>% 
  filter(uptake > 30)
```

But we get an error, indicating that the "uptake" object cannot be
found. Why is that the case? From a programming perspective, it's
possible for multiple functions with the same name to coexist, and when
loading them collectively, certain function names may get overridden. In
this particular case, we wanted to use the `filter()`function from the
`{dplyr}` package but instead are using the `filter()`function from the
`{stats}` package. We can verify this checking the function
documentation (by placing the cursor on the `filter()`function and
pressing `F1`), as seen in the image below.

![Checking the `filter()` function from the stats
packages.](images/filter_stats.png){#image-filter-stats align="center"}

This can frequently occur when you have multiple packages loaded, and
you may not always ensure that the functions are being overwritten or
not. This is much more likely to happen when you load a lot of packages
and don't check to make sure that the functions you are using are
overwritten or not.

When we load the `{dplyr}` package and then use the `filter()`function,
it finally works:

```{r}
#| error: true
#| message: false
library(dplyr)
CO2 %>% 
  filter(uptake > 30)
```

### How to prevent Namespace issues?

The following function can be used as a tool to find conflicts with any
function from all the packages you have loaded in your session:

```{r}
conflicted::conflict_scout()
```

Thanks to this function we can visualize that there are two possible
conflicts with the `filter()`and `lag()` functions, which exist in both
packages `{dplyr}` and `{stats}`.

Another good practice is to specify the packages in the functions like
follows:

```{r}
#| eval: false
stats::filter()
```

In the previous case, we specify that we want to use the `filter`
function in the `{stats}` package.

## Reload, reinstall packages, and start R session

The following error informs us that one the packages is needing a higher
version of another package than the one is currently loaded:

![Different package version
errors.](images/dplyr_error.png){#image-version-error align="center"}

To solve this issue, you can try to:

1.  Re-nstall packages. The following code string will force the
    re-installation of packages:

    ```{r}
    #| eval: false
    install.packages("rlang", force = TRUE)
    ```

2.  Load the packages one by one.

3.  Restart the R session if all the packages are correctly installed
    and loaded (image below).

![Restart the R
session.](images/restart_rsession.png){#image-restart-rsession
align="center"}

## Debug functions

In this section we cover the functions `debugonce()` and `browser()`.

The function `debugonce()` allows us to set a one-time breakpoint in our
code. When entering `debugonce(function_name)`, it modifies the function
so that when we run the function next, it causes it to enter into
debugging mode. This allows us to interactively debug and inspect the
function. Let's explore an example involving a custom filtering function
that contains an issue:

```{r}
#| error: true
subset_co2 <- function(filter_condition) {
  dplyr::filter(CO2, filter_condition)
}
subset_co2(uptake > 35)
```

As we see from the error message, the problem seems to be that the
object `uptake` is not found, even though we know that if we used that
same code while using `filter()`, it would work. We can run
`debugonce()` on the function and it will enter debug mode:

```{r}
#| eval: false
debugonce(subset_co2)
subset_co2(uptake > 35)
```

This mode is shown in the image below. Using the "next", "continue", and
"stop" buttons that are found just above the RStudio Console, we can
sequentially navigate line by line through your function and check the
objects in our environment (the A section of the image).

A similar function is the `browser()`. This function will cause the
function to enter the debug state at the line it is included in. So, you
can place it where you think your problem is and start debugging from
there. We run the following code:

```{r}
#| error: true
#| eval: false
subset_co2 <- function(filter_condition) {
  browser()
  dplyr::filter(CO2, filter_condition)
}
subset_co2(uptake > 35)
```

We re-enter the debugging state again. If we type out `filter_condition`
into the Console (the B panel in the image), we see that it is not
properly entering into the function. So the problem doesn't come from
within the function itself, the issue lies in the code outside the
function.

R is a programming language created for data analysis and statistics
rather than computer science. [Non-standard
evaluation](https://adv-r.hadley.nz/quasiquotation.html#quasi-motivation)
refers to the code that is evaluated dynamically by expressions rather
than by its value, allowing flexibility and a dynamic behavior. It is
really useful when you are writing statistical code, but it can
potentially introduce a cascade of errors when used in situations that
require standard evaluation in the code strings.

Non-standard evaluation can be better explained by code that is quoted
or unquoted. A very common example is when you use `library()`:

```{r}
# Non-standard evaluation
library(dplyr)

# Standard evaluation
library("dplyr")
```

The function `library()` uses both standard and non-standard evaluation
here. It's called non-standard because normally when code is unquoted
like with `dplyr`, programming languages evaluate it by running it. But
we know that `dplyr` isn't actually an object in R so can't run:

```{r}
#| error: true
dplyr
```

But when quoted:

```{r}
"dplyr"
```

It gives us a character value back. R in general but especially
`{tidyverse}` use non-standard evaluation frequently, since it makes
coding and reading code easier, but makes errors and programming a bit
more difficult.

So with our filtering condition (`uptake > 30`), we assumed we were
using non-standard evaluation. But only some functions are designed for
that, and our custom function is not designed to use it. So when we run
it, R looks for the object `uptake`, but it doesn't exist. It is a
column in the dataset, but not an object in R.

Thankfully, `{tidyverse}` provides a method to easily use non-standard
evaluation within custom functions. This method is called curly-curly
(`{{ }}`), which we can use within the `filter()` function in our custom
function (see in panel C of the image). This revised code below will
work without issues.

```{r}
subset_co2 <- function(filter_condition) {
  dplyr::filter(CO2, {{ filter_condition }})
}
subset_co2(uptake > 35)
```

![Debugging functions with `browser()` and
`debugonce()`.](images/debugging_functions.png){align="center"}
